---
title: "Yelp Data Wrangling Project"
author: "Rich Dean"
date: "22 October 2015"
output: pdf_document
---


This document covers my initial exploration of the Yelp Dataset Challenge data. This dataset from the online review site Yelp, and is part of a data anlysis competition, now in its fifth run.
The data contains (from the Yelp website):

* 1.6M reviews and 500K tips by 366K users for 61K businesses
* 481K business attributes, e.g., hours, parking availability, ambience.
* Social network of 366K users for a total of 2.9M social edges.
* Aggregated check-ins over time for each of the 61K businesses

### Initialisation
To start, I'll import my libraries, include files, and set globals variables. After that, I'll load in the data (from cache if available).
```{r initialise, results='hide', message=FALSE, warning=FALSE}

library(jsonlite)
library(ggmap)
library(ggplot2)
library(GGally)
library(RColorBrewer)
library(tidyr)
library(dplyr)
# Include my standard function library
source(file="~/R_code/common/rrfuncs.R")
# Optionally clear my environment - makes things clearer
# rm(list=ls())
data_dir = c('~/datasets/yelp/')
```
```{r load_data, cache=TRUE}
if (file.exists(paste0(data_dir,'yelp_data.Rd')))
{
  load(paste0(data_dir,'yelp_data.Rd'))
} else {
  # The data is in a line-by-line JSON format, rather than fully-compliant JSON. 
  # Use 'stream_in' from jsonlite
  y.business <- stream_in(file(paste0(data_dir,
                                      "yelp_academic_dataset_business.json")))
  y.checkin <- stream_in(file(paste0(data_dir,
                                     "yelp_academic_dataset_checkin.json")))
  y.review <- stream_in(file(paste0(data_dir,
                                    "yelp_academic_dataset_review.json")))
  y.tip <- stream_in(file(paste0(data_dir,"yelp_academic_dataset_tip.json")))
  y.user <- stream_in(file(paste0(data_dir,"yelp_academic_dataset_user.json")))
  # Cache the R data
  save(list=c('y.business','y.user','y.review','y.tip','y.checkin'), 
       file=paste0(data_dir,'yelp_data.Rd'))
}
```

Now that the data is loaded, I'll quickly verify the Yelp statements about the dataset size, and take a quick look at the structure of the **y.business** data frame.
```{r datacheck}
data.frame(rbind(
    c(Stat='Reviews',Rows=nrow(y.review)), 
    c('Tips', nrow(y.tip)),
    c('Users', nrow(y.user)),
    c('Businesses', nrow(y.business)))
  )
# Use max.level, as this is a coplex nested frame
str(y.business, max.level=1)
```

## Geography

I know that the businesses are located in 10 distinct areas. I have their names, and the **latitude**/**longitude** are in the **y.business** object.
I'll use k-means clustering to label the groups, which will allow for some comparative geo-analysis. I have the list of actual locations which the data should be from, so i will seed the kmeans with the lat/long of those ten locations.
```{r geocode,cache=TRUE, message=FALSE, warning=FALSE}
cities<-c('Edinburgh, UK', 'Karlsruhe, Germany', 'Montreal, Canada', 
          'Waterloo, Canada', 'Pittsburgh, PA', 'Charlotte, NC', 
          'Urbana-Champaign, IL', 'Phoenix, AZ', 'Las Vegas, NV', 'Madison, WI')
region.centres<-geocode(cities, source='google')
set.seed(43046721)
myclus<-kmeans(y.business[,c('longitude','latitude')],region.centres)
y.business$location<-cities[myclus$cluster]
table(y.business$location, y.business$state)
ggplot(data=y.business, aes(x=location)) + 
  geom_histogram() +
  theme(axis.text.x=element_text(angle=50, size=10, vjust=0.5))
```

These results show a good grouping, where each state is entirely represented by one cluster, with the exception of two. 
Cluster 8 and 9 both contain CA - reasonable as both *AZ* and *NV* border it. 
Cluster 9 seems to contain a rogue *SC* entry:
```{r}
subset(y.business, state=='NC' & location=='Las Vegas, NV')$full_address
```

It's a typo! Yelp must take the **state** value from the entered address. 
I'll do a pair of map plots - one for the US, and one for Europe - to show the distribution of the business, coloured by the cluster.
```{r create_maps,message=FALSE, error=FALSE}
europe.centre <- colMeans(apply(myclus$centers[cities %in% 
                            c('Edinburgh, UK','Karlsruhe, Germany'),],2,range))
map.europe <- get_map(location = europe.centre, zoom = 6)
map1<-ggmap(map.europe) + 
    geom_point(aes(x=longitude, y=latitude, color=location), 
               data=subset(y.business, 
                        location %in% c('Edinburgh, UK','Karlsruhe, Germany')))

usa.centre <- colMeans(apply(myclus$centers[!cities %in% 
                            c('Edinburgh, UK','Karlsruhe, Germany'),],2,range))
map.usa <- get_map(location = usa.centre, zoom = 4)
map2<-ggmap(map.usa) + 
    geom_point(aes(x=longitude, y=latitude, color=location), 
               data=subset(y.business, 
                        !location %in% c('Edinburgh, UK','Karlsruhe, Germany')))

map1
map2
```

The maps show that the clustering worked nicely, and also shows the geographic spread of the businesses in the dataset. Despite the scale of the USA map, you can see something of the dispersal of businesses within each cluster - the Phoenix AZ cluster is the largest of the groups on the map, while the Urbana-Champaign IL group is smallest.

I'll add a distance to the cluster center too - I can then do some analysis of the area covered by each cluster
This is only for a comparative measure, and the distances /should/ be relatively short, so i won't go so far as a great circle calculation
```{r geo_distance}
t.distance<-y.business[,c('longitude','latitude')]-myclus$centers[
                                                                myclus$cluster,]
y.business$distance.from.centre<-sqrt(rowSums(t.distance**2))
tbl_df(y.business[,c('location','distance.from.centre')]) %>% 
  group_by(location) %>% 
  summarise(mn=mean(distance.from.centre), sd=sd(distance.from.centre))
```
This shows that Urbana-Champaign has the tightest group of businesses (low mean and standard deviation), while Phoenix has the most dispersed, as seen on the map plots.

****************************************  TODO   MORE GEO STUFF!!!!!   Maybe a multivariate plot of some kind? Deeper analysis of a single city?

## Business Categories

I'd also like to get a primary category for each business - this will also simplify analysis. The primary category list is taken from https://www.yelp.com/developers/documentation/v2/all_category_list.

```{r primary_category, cache=TRUE}
yelp.primary.categories<-
  c('Active Life','Arts & Entertainment','Automotive','Beauty & Spas',
    'Education','Event Planning & Services','Financial Services','Food',
    'Health & Medical','Home Services','Hotels & Travel','Local Flavor',
    'Local Services','Mass Media','Nightlife','Pets','Professional Services',
    'Public Services & Government','Real Estate','Religious Organizations',
    'Restaurants','Shopping')

y.business$primary.categories<-
  lapply(y.business$categories, 
         function (x) sort(intersect(x, yelp.primary.categories)))
length(unique(y.business$primary.categories))
```
There are 325 combinations of primary categories! This makes any specific comparison of primary categories tricky, as businesses could belong to several primary categories. Taking an slightly different tack led me to think more deeply about the categories and review counts. Do businesses with more categories have more reviews and/or higher ratings?

```{r category_plot}
cat.dist<-
  data.frame(categories=
               as.factor(
                 sapply(y.business$categories, 
                        function(x) length(unlist(x)))),
             location=y.business$location,
             reviews=y.business$review_count, 
             stars=y.business$stars)
# Build a second variable to hold summaries
cat.dist2<- cat.dist %>% 
            group_by(categories) %>% 
            summarise(median_reviews=median(reviews), 
                      median_stars=median(stars), 
                      n=n())
rr.multiplot(
  ggplot(data=cat.dist2, aes(x=categories)) + 
    geom_histogram(aes(y=median_reviews), stat='identity'),
  ggplot(data=cat.dist2, aes(x=categories)) + 
    geom_histogram(aes(y=median_stars), stat='identity'),
  cols=2)
```

There isn't a clear link between the ratings and number of categories, but there appears to be a clear link between categories and number of reviews

```{r plot_ratings_reviews}
ggplot(data=cat.dist, aes(x=categories, y=reviews)) +
  coord_trans(ytrans='log10') +
  geom_boxplot() +
  geom_jitter(alpha=0.2, aes(color=categories), 
              position=position_jitter(width=.2))+
  scale_colour_brewer('categories', palette='Paired') +
  scale_y_log10()
```

Using a log plot of reviews shows there are actually few businesses with very high numbers of reviews. The point density shows the bulk of businesses have between 2 and 5 categories, and less than 100 reviews


## Checkins

I had hoped to perform more category-based analysis, but with multiple primary categories, I would have to sift the combinations and derive so kind of hierarchy to determine a single primary for each business. Instead, I chose to look into a different view of the data - checkins. This shows when businesses are visited, by week-hour. The **y.checkin** data frame contains my core data - I'll add some of the columns from **y.business** to provide a more useful single frame. I will use heatmaps to visualise this data.

```{r checkin_build,cache=TRUE}
y.checkin<-data.frame(flatten(y.checkin))
y.checkin<- 
  y.checkin %>%
  left_join(y.business[,c('business_id','location','review_count',
                          'stars', 'primary.categories')], 
            by = 'business_id')
# Need to gather the checkin_info columns
# I'll split the check-in column names to give hour/day values
y.checkin.long <- 
  y.checkin %>%
  gather(key, value, -location, -stars, -review_count, -type, -business_id, 
         -primary.categories, na.rm=TRUE) %>%
  mutate(checkin.list= strsplit(as.character(key), '\\.')) %>%
  mutate(hr = as.integer(sapply(checkin.list, function (x) x[2])),
         dy = sapply(checkin.list, function (x) x[3])) %>%
  select(-key, -checkin.list, -type, business_id) 

# Break the primary category list into columns
# This code from http://stackoverflow.com/questions/25347739/r-convert- \\ 
#                     factor-column-to-multiple-boolean-columns
lvl <- unique(unlist(y.checkin.long$primary.categories))      
res <- data.frame(do.call(rbind,lapply(y.checkin.long$primary.categories,
                                       function(x) table(factor(x, levels=lvl)))
                          ), 
                  stringsAsFactors=FALSE)
y.checkin.long<-cbind(y.checkin.long, res)


#  Lets look at restaurants, nightlife, and religious organisations
rr.heatmap <- function (l.criteria, loc_wrap = FALSE)
{
  p<-ggplot(subset(y.checkin.long, eval(parse(text=l.criteria))), aes(dy, hr)) + 
    geom_tile(aes(fill = log(value)), colour = "white") + 
    scale_fill_gradient(low = "white", high = "red")
  p + if (loc_wrap==TRUE) facet_wrap( ~ location ) else NULL
}

rr.heatmap('Restaurants>0&value<quantile(value,0.95)')+ggtitle('Restaurants')
``` 

I would have expected to see a pattern in restaurant checkins - more frequent at lunchtimes and evenings - but there are no clear patterns to be seen in this plot. The *nightlife* category should clearly show a pattern...
```{r checkin_heatmap2}
rr.heatmap('Nightlife>0&value<quantile(value,0.95)')+ggtitle('Nightlife')
```
This very puzzling. The checkins seem to be widely scattered. A final look at these, taking Religious Organisations and splitting by location...
```{r checkin_heatmap3}
rr.heatmap('Religious.Organizations>0&value<quantile(value,0.95)', TRUE)+
  ggtitle('Religious Organisations')
```
This plot does show a few patterns. Phoenix has a clear grouping of checkins on day 6, around 10am - morning on a Sunday seems a sensible time for visitng church. The Las Vegas distribution is clearly weighted towards the evening, and Saturday morning and lunchtime. My theory here would be that the checkins are mainly at wedding venues. Sadly the dataset doesn't contain business names, so this would be difficult to investigate further. It would appear that checkins are not always associated with actual presence at the business in question, which makes the data of limited real use.

## Users

******************  TODO Expand on the text 
Initial look at the user data
```{r user_plot1}
y.user$elite.years<-unlist(lapply(y.user$elite, length))

# Sample 50k rows to increase speed 
ggplot(data=sample_n(y.user,50000), aes(x=review_count, y=fans)) + 
  geom_point( alpha=0.2, position='jitter')

```
Improve - remove extreme outliers, log the scales, add the elite.years as a colour 
```{r user_plot2}

ggplot(data=
       subset(y.user, 
              review_count<quantile(review_count,0.99) 
              & fans<quantile(fans,0.99)
              ), 
       aes(x=review_count, y=fans)
       ) + 
  geom_point(alpha=0.2, 
             position='jitter', 
             aes(color=elite.years+0.1)
             )+
  scale_colour_gradientn(
             colours=c('blue','green'), 
             trans='log', 
             breaks=c(1,3,5,10,15))
```
****** TODO Correlation between review count and fans
****** TODO Further user analysis - 2-3 more vis

## Final Plots and Summary


****** TODO Select my 3 plots, make them perfect. Suggest:
****** TODO USA map, heat map of religion in Phoenix, one other TBD

## Reflections

******* TODO 200 words
Large multi-layered dataset. 
Primarily categorical/ text based. Well suited for a NLP/prediction project.



